---
title: "Getting started with digitraffic"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with digitraffic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## What is Digitraffic?

[Digitraffic](https://www.digitraffic.fi/) is an open API maintained by
Finntraffic that provides real-time and historical traffic data from Finland's
road network. The data comes from over 450 **LAM** (_Liikenteen Automaattinen
Mittaus_ / Automatic Traffic Measurement) stations — electromagnetic induction
loops embedded in the road surface that detect every passing vehicle.

Each LAM station records:

- **Speed** (km/h)
- **Vehicle length** (metres) — used for classification
- **Direction and lane**
- **Headway** — time gap to the previous vehicle
- **Vehicle classification** — 7 categories from cars to HCT trucks

The `digitraffic` R package wraps these API endpoints and returns all data as
tidy tibbles, ready for analysis with dplyr and visualisation with ggplot2.

## Installation

```{r install}
# install.packages("pak")
pak::pak("theo-blauberg/digitraffic")
```

## Loading the package

```{r setup}
library(digitraffic)
library(dplyr)
```

## Discovering stations

The first thing you'll want to do is find stations of interest. `dt_stations()`
returns a tibble of all LAM stations with their coordinates, bearing, and
operational status:
```{r stations}
stations <- dt_stations()
stations
```

The result is cached for 5 minutes so repeated calls are instant.
`dt_stations()` supports five filter parameters — see
`vignette("station-filtering")` for the full guide.  Quick examples:

```{r stations-filter}
# By name (case-insensitive regex)
dt_stations(name = "Espoo")

# By road number (parsed from name, no extra API calls)
dt_stations(road_number = 1)

# By geographic bounding box
dt_stations(bbox = c(24.5, 60.1, 25.2, 60.4))

# By municipality or province (uses bundled metadata cache)
dt_stations(municipality = "Espoo")
dt_stations(province = "Uusimaa")

# Filters combine — road 4 stations in Pirkanmaa
dt_stations(road_number = 4, province = "Pirkanmaa")
```

### Detailed station metadata

For a single station, `dt_station()` returns an extended set of 32 fields
including municipality, province, road number, road section, free-flow
speeds, and multilingual names:

```{r station-detail}
detail <- dt_station(23001)
detail

# Key fields
detail |>
  select(name, municipality, province, road_number,
         free_flow_speed1, free_flow_speed2)
```

### Mapping stations with sf

If you have the `sf` package installed, you can convert the stations tibble
to a spatial object for mapping:

```{r sf-stations}
library(sf)

stations_sf <- dt_stations() |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

# Filter to active stations only
active <- stations_sf |>
  filter(collection_status == "GATHERING")

plot(active["bearing"], pch = 20, main = "LAM stations coloured by bearing")
```

## Real-time sensor data

`dt_station_data()` fetches the current sensor readings for one station.
Data is updated approximately every minute:

```{r sensor-data}
data <- dt_station_data(23001)
data
```

Each row is a single sensor measurement. The `name` column contains a
structured Finnish name encoding the measurement type, interval, and
direction. For example:

| Sensor name | Meaning |
|---|---|
| `OHITUKSET_5MIN_KIINTEA_SUUNTA1` | Vehicle count, 5-min fixed interval, direction 1 |
| `KESKINOPEUS_60MIN_KIINTEA_SUUNTA1` | Average speed, 60-min fixed interval, direction 1 |
| `OHITUKSET_5MIN_LIUKUVA_SUUNTA2_KAISTA1` | Vehicle count, 5-min sliding, direction 2, lane 1 |

### Filtering to speed readings

```{r speed-readings}
data |>
  filter(unit == "km/h") |>
  select(name, value, unit, time_window_start, time_window_end)
```

### Filtering to volume readings

```{r volume-readings}
data |>
  filter(unit == "kpl/h") |>
  select(name, value, unit, time_window_start, time_window_end)
```

### All stations at once

`dt_stations_data()` fetches sensor data for every station in one request.
The response is several MB, so it takes a few seconds:

```{r all-stations-data}
all_data <- dt_stations_data()

# How many stations have data?
n_distinct(all_data$station_id)

# Average 60-min speed across all stations, direction 1
all_data |>
  filter(name == "KESKINOPEUS_60MIN_KIINTEA_SUUNTA1") |>
  summarise(
    n_stations    = n(),
    mean_speed    = mean(value, na.rm = TRUE),
    median_speed  = median(value, na.rm = TRUE),
    min_speed     = min(value, na.rm = TRUE),
    max_speed     = max(value, na.rm = TRUE)
  )
```

## Understanding sensor types

There are 76 different sensor types. `dt_sensors()` returns the full
catalogue with descriptions:

```{r sensors}
sensors <- dt_sensors()
sensors

# Speed sensors
sensors |> filter(unit == "km/h")

# Volume sensors
sensors |> filter(unit == "kpl/h")

# Status / special sensors
sensors |> filter(unit == "***")
```

You can join this catalogue to sensor data to enrich the output:

```{r join-sensors}
dt_station_data(23001) |>
  left_join(
    sensors |> select(sensor_id = id, description_fi, direction),
    by = "sensor_id"
  )
```

## Sensor constants (calibration values)

Each station has calibration constants that define reference values for that
specific road segment:

```{r constants}
# All constants
consts <- dt_sensor_constants()
consts

# Constants for one station
dt_sensor_constants(id = 23001)
```

The constants include:

| Name | Meaning |
|---|---|
| `VVAPAAS1` | Free-flow speed (km/h), direction 1 |
| `VVAPAAS2` | Free-flow speed (km/h), direction 2 |
| `MS1` | Maximum hourly capacity (vehicles/h), direction 1 |
| `MS2` | Maximum hourly capacity (vehicles/h), direction 2 |
| `Tien_suunta` | Road bearing in degrees (0-360) |

Many stations have **seasonal variants** with different summer (Apr-Oct) and
winter (Nov-Mar) values:

```{r seasonal-constants}
dt_sensor_constants(id = 20002) |>
  filter(grepl("VVAPAAS", name)) |>
  select(name, value, valid_from, valid_to)
#> # A tibble: 4 x 4
#>   name     value valid_from valid_to
#>   <chr>    <dbl> <chr>      <chr>
#> 1 VVAPAAS1   105 04-01      10-31
#> 2 VVAPAAS1   100 11-01      03-31
#> 3 VVAPAAS2   105 04-01      10-31
#> 4 VVAPAAS2   100 11-01      03-31
```

## Next steps

- For filtering stations by road, municipality, province, and bounding box, see `vignette("station-filtering")`
- For historical per-vehicle data, see `vignette("historical-data")`
- For the full function reference, see the [package documentation site](https://theo-blauberg.github.io/digitraffic/)
