---
title: "Filtering and discovering stations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Filtering and discovering stations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

Finland's LAM network has over 500 measurement stations spread across the road
network.  `dt_stations()` supports five complementary filter parameters so you
can narrow down to exactly the stations you need without writing manual
`dplyr::filter()` calls.

| Parameter | Filters by | Data source | Extra API calls? |
|---|---|---|---|
| `name` | Station name (regex) | Bulk endpoint | None |
| `road_number` | Road number (e.g. 7 for vt7) | Parsed from name | None |
| `bbox` | Geographic bounding box | Bulk coordinates | None |
| `municipality` | Municipality name | Detailed cache | None |
| `province` | Province name | Detailed cache | None |

All filters are combinable and applied in the order listed above.

```{r setup}
library(digitraffic)
library(dplyr)
```

## Filter by name

The simplest filter — a case-insensitive regular expression matched against
the station name string.  Station names encode road type, road number, and
location, e.g. `"vt1_Espoo_Hirvisuo"`:

```{r name-filter}
# Stations whose name contains "Espoo" (case-insensitive)
dt_stations(name = "Espoo")

# Exact road prefix — valtatie 1 only
dt_stations(name = "^vt1_")

# All kantatie stations
dt_stations(name = "^kt")

# Helsinki or Espoo in the name
dt_stations(name = "Helsinki|Espoo")
```

## Filter by road number

`road_number` extracts the integer road number from the station name string —
no extra API calls required.  The naming convention covers all Finnish road
types:

| Name prefix | Road type | Example |
|---|---|---|
| `vt` | Valtatie (national road) | `vt7_Rita` → 7 |
| `kt` | Kantatie (primary road) | `kt51_Keha` → 51 |
| `st` | Seututie (regional road) | `st140_Sipoo` → 140 |
| `mt` | Maantie (local road) | `mt120_Muu` → 120 |

```{r road-filter}
# All stations on valtatie 7 (Helsinki–Russian border)
dt_stations(road_number = 7)

# All stations on the Ring Road III (kt50)
dt_stations(road_number = 50)

# Stations on roads 1 through 5
lapply(1:5, dt_stations, road_number = ) |>
  dplyr::bind_rows(.id = "road")
```

Stations whose names don't follow the standard convention return `NA` for the
parsed road number and are silently excluded from `road_number` filter results.

## Filter by bounding box

Pass a length-4 numeric vector `c(lon_min, lat_min, lon_max, lat_max)` in
WGS-84 degrees.  Works on the bulk coordinates already returned by
`dt_stations()` — no extra API calls:

```{r bbox-filter}
# Helsinki metropolitan area
dt_stations(bbox = c(24.5, 60.1, 25.2, 60.4))

# Greater Tampere region
dt_stations(bbox = c(23.4, 61.3, 24.1, 61.7))

# Entire southern Finland
dt_stations(bbox = c(22.0, 59.8, 27.0, 61.5))
```

You can combine `bbox` with `road_number` to find stations on a specific road
within a geographic area:

```{r bbox-road}
# valtatie 4 stations south of Lahti
dt_stations(road_number = 4, bbox = c(25.5, 60.9, 26.5, 61.2))
```

### Getting coordinates for a bbox

If you have the `sf` package, you can compute a bounding box from a polygon:

```{r sf-bbox}
library(sf)

# Example: use a municipality polygon to build a bbox
# espoo_polygon <- ... (loaded from a shapefile or administrative boundary API)
# bb <- st_bbox(espoo_polygon)
# dt_stations(bbox = c(bb["xmin"], bb["ymin"], bb["xmax"], bb["ymax"]))
```

## Filter by municipality or province

These filters use **detailed station metadata** — 32 fields per station
including municipality, province, road address, and multilingual names.  This
data is served from a cache so no API calls are made at filter time.

```{r municipality-filter}
# All stations in Espoo
dt_stations(municipality = "Espoo")

# Stations in Tampere (case-insensitive)
dt_stations(municipality = "tampere")

# All stations in Helsinki or Espoo (regex)
dt_stations(municipality = "Helsinki|Espoo")
```

```{r province-filter}
# All stations in Pirkanmaa province
dt_stations(province = "Pirkanmaa")

# Stations in either of two provinces
dt_stations(province = "Uusimaa|Pirkanmaa")
```

Combine with other filters freely:

```{r combined-filters}
# valtatie 3 stations in Pirkanmaa
dt_stations(road_number = 3, province = "Pirkanmaa")

# Stations in Espoo within a tight bounding box
dt_stations(municipality = "Espoo", bbox = c(24.5, 60.1, 25.0, 60.4))
```

## The detailed station cache

Municipality and province filtering relies on a **detailed metadata cache**
that holds 32 fields per station (vs. the 10 fields from the bulk endpoint).

### What the cache contains

The cache is built by fetching the individual station endpoint
(`/api/tms/v1/stations/{id}`) for every station.  It includes:

- `municipality` / `municipality_code`
- `province` / `province_code`
- `road_number` / `road_section` / `distance_from_section_start`
- `direction1_municipality` / `direction2_municipality`
- `free_flow_speed1` / `free_flow_speed2`
- `name_fi` / `name_sv` / `name_en`
- and more — see `?dt_station` for the full list

### Two-tier cache system

```
Priority 1 (user disk):  tools::R_user_dir("digitraffic", "cache")/stations_detailed.rds
                                  |
                         (falls back if not present)
                                  |
Priority 2 (bundled):    R/sysdata.rda  (baked into the package at build time)
```

The bundled snapshot is always present — `dt_stations(municipality = "Espoo")`
works immediately after installing the package, with no setup required.

### Integrity check

Each time you use a detail-based filter, the package compares the cached
station IDs against the current live station list from `dt_stations()`.  If
they differ (new stations added, old ones decommissioned), you'll see:

```
Warning: Detailed station cache is out of date: 3 new stations detected.
i Run `dt_stations_load_details()` to refresh.
```

Your filter still runs — it just uses the slightly stale cache.

### Refreshing the cache

When the network has changed, run:

```{r load-details}
dt_stations_load_details()
#> Fetching detailed metadata for 521 stations...
#> [working] ■■■■■■■■■■■■■■■■■■■■■■■■■■■■■■■  100% | ETA:  0s
#> v Saved detailed metadata for 521 stations.
#> i Cache location: ~/.cache/R/digitraffic/stations_detailed.rds
```

This uses `httr2::req_perform_parallel()` with 10 concurrent connections.
On a typical connection it completes in under 15 seconds.  The result is saved
to your personal cache directory and persists across R sessions.

To control the concurrency:

```{r load-details-concurrency}
# More aggressive (faster on a fast connection)
dt_stations_load_details(max_active = 20)

# More conservative (politer to the API)
dt_stations_load_details(max_active = 5)
```

## Using filtered results with sensor data

A common pattern: filter stations first, then fetch sensor data for each:

```{r filter-then-fetch}
# Get all stations in Uusimaa
uusimaa <- dt_stations(province = "Uusimaa")

# Fetch real-time data for the first 5
sensor_data <- lapply(head(uusimaa$id, 5), dt_station_data) |>
  dplyr::bind_rows()

# Average speed in each direction
sensor_data |>
  filter(grepl("KESKINOPEUS_60MIN.*SUUNTA", name)) |>
  group_by(name) |>
  summarise(mean_speed = mean(value, na.rm = TRUE))
```

## Mapping filtered stations

Combine filtering with `sf` for spatial analysis and mapping:

```{r sf-mapping}
library(sf)
library(ggplot2)

# Stations in Pirkanmaa projected to Finnish TM35FIN (EPSG 3067)
pirkanmaa_sf <- dt_stations(province = "Pirkanmaa") |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(3067)

# Plot with ggplot2
ggplot(pirkanmaa_sf) +
  geom_sf(aes(colour = collection_status), size = 2) +
  scale_colour_manual(values = c(
    "GATHERING"           = "#2166ac",
    "REMOVED_TEMPORARILY" = "#d73027"
  )) +
  labs(
    title  = "LAM stations in Pirkanmaa",
    colour = "Status"
  ) +
  theme_minimal()
```

## Summary of filter behaviour

```{r summary-table, eval=FALSE}
# Quick reference — all these are equivalent to chaining dplyr::filter() calls
# but done in a single function with no repeated API calls

dt_stations()                                    # all 500+ stations
dt_stations(name = "^vt1_")                      # name regex
dt_stations(road_number = 1)                     # road number from name
dt_stations(bbox = c(24.5, 60.1, 25.2, 60.4))   # bounding box
dt_stations(municipality = "Espoo")              # municipality (cache)
dt_stations(province = "Uusimaa")                # province (cache)

# Combined — all filters applied in sequence
dt_stations(
  road_number  = 1,
  province     = "Uusimaa",
  bbox         = c(24.5, 60.1, 25.2, 60.4)
)
```

## See also

- `dt_station()` — detailed metadata for a single station (all 32 fields)
- `dt_stations_load_details()` — refresh the detailed metadata cache
- `vignette("digitraffic")` — getting started with real-time sensor data
- `vignette("historical-data")` — historical per-vehicle analysis
